\textit{MSF: Modulated Sub-graph Finder}

### 2018-08-09 MSF_Paper_FA.tex
Supplement ref

Existing pathway-based analysis approaches use different research
designs, which can be categorized into ORA (Over-representation
analysis), FCS (Functional class scoring) and pathway topology based
methods. All of which aim to find a subset of genes, e.g.,
significantly differentially expressed genes, genes associated with a
certain pathway more often than expected given the total set of
examined genes, e.g.~the whole genome background.  \ILH{ORA \cite{} is
  the first and the most basic method of pathway analysis.} It uses a
DEG list with user defined cut-off for the log-fold change and
\textit{p}-value (most commonly using absolute log-fold change $\geq$
2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
associated with annotated pathways are tested for being
over-represented in the set of DEGs. To this end, hyper-geometric
distribution, chi-square tests, binomial probability or the Fisher’s
exact test are used. Thereby the information of the topology of genes
in the pathways is neglected~\cite{Bayer}. Furthermore, ORA assumes
that the biological pathways are independent of each other and ignores
the fact that they cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

In the next step, we check if any of the initial modulated sub-graphs 
could further be extended by adding more than one gene at a time. This is done
by testing all possible extension paths up to \emph{N} (usually 2) genes at
all nodes in the sub-graph. Again, this step is iteratively repeated until
no further genes are added to the significant differentially expressed
sub-graphs. This step bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

@@

After detection and extension of the modulated sub-graphs, they are tested
if combined sub-graphs score better than on their own. The merging of the
two sub-graphs is done by depth first search \ILH{unclear}. If the two sub-graphs merge
with the connector of at most \emph{N} genes (default two genes) and the
combined \textit{p}-value of the merged sub-graph including the bridging
genes in between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

@@

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, three large modulated sub-graphs were identified with 41,
107, and 18 genes were identified, as well as two with less than 4 genes.
The modulated sub-graphs consist predominantly of cytokines, chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23).
IFNB1 and IFNA1 were both identified as two of the possible sources
\ILH{(in all subgraphs?)}.
Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes \ILH{(link to some supplementary
  material with identified sub-graphs)}.
At 24~hpi nine modulated sub-graphs were
identified with four main sub-graphs consisting
of 38, 85, 148 and 167 genes, others being smaller than 10 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point 48~hpi,
eleven modulated sub-graphs were identified. Eight of the sub-graphs were
less than seven genes and main sub-graphs had 39, 71 and 210 genes. IFNB1
and IFNA1 were identified as \TODO{the two sources out of the total sources}.

@@

\noindent
Source code:
https://github.com/MariamFarman/Modulated-SubGraph-Finder \newline Software
license: Creative Commons Attribution 4.0 International License

@@

\begin{table*}[]
	\centering
	\caption{Comparison of connected sub-graphs of modulated genes
          in the global network between the analysis results of
          \texttt{MSF} and mapping the raw list of differentially
          expressed genes from the standard edgeR analysis, applying
          different p-value cut-offs, onto the network. }
	\label{tab:rawVsHartung}
	\begin{tabular}{lll}
		\hline
		\multicolumn{1}{|l|}{}                            & \multicolumn{1}{c|}{\textbf{Total number of}}       & \multicolumn{1}{c|}{\textbf{Number of connected}}      \\
                \multicolumn{1}{|l|}{}                            & \multicolumn{1}{c|}{\textbf{genes in network}}      & \multicolumn{1}{c|}{\textbf{sub-graph in network}}     \\ \hline
		\multicolumn{1}{|l|}{\textbf{6hpi}}               & \multicolumn{1}{l|}{}                                                & \multicolumn{1}{c|}{}                 \\ \hline
		\multicolumn{1}{|l|}{edgeR + MSF}                 & \multicolumn{1}{c|}{166}                                             & \multicolumn{1}{c|}{3}                \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.1)}  & \multicolumn{1}{c|}{123}                                             & \multicolumn{1}{c|}{39}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.05)} & \multicolumn{1}{c|}{123}                                             & \multicolumn{1}{c|}{47}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.01)} & \multicolumn{1}{c|}{112}                                             & \multicolumn{1}{c|}{53}               \\ \hline
		\multicolumn{1}{|l|}{\textbf{24hpi}}              & \multicolumn{1}{c|}{}                                                & \multicolumn{1}{c|}{}                 \\ \hline
		\multicolumn{1}{|l|}{edgeR + MSF}                 & \multicolumn{1}{c|}{438}                                             & \multicolumn{1}{c|}{4}                \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.1)}  & \multicolumn{1}{c|}{325}                                             & \multicolumn{1}{c|}{76}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.05)} & \multicolumn{1}{c|}{305}                                             & \multicolumn{1}{c|}{102}              \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.01)} & \multicolumn{1}{c|}{262}                                             & \multicolumn{1}{c|}{110}              \\ \hline
		\multicolumn{1}{|l|}{\textbf{48hpi}}              & \multicolumn{1}{c|}{}                                                & \multicolumn{1}{c|}{}                 \\ \hline
		\multicolumn{1}{|l|}{edgeR + MSF}                 & \multicolumn{1}{c|}{320}                                             & \multicolumn{1}{c|}{3}                \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.1)}  & \multicolumn{1}{c|}{226}                                             & \multicolumn{1}{c|}{39}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.05)} & \multicolumn{1}{c|}{207}                                             & \multicolumn{1}{c|}{62}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.01)} & \multicolumn{1}{c|}{173}                                             & \multicolumn{1}{c|}{67}               \\ \hline
	\end{tabular}
\end{table*}

>>>>

Existing pathway-based analysis approaches use different research
designs, which can be categorized into ORA (Over-representation
analysis), FCS (Functional class scoring) and pathway topology based
methods. All of which aim to find a subset of genes, e.g.,
significantly differentially expressed genes, genes associated with a
certain pathway more often than expected given the total set of
examined genes, e.g.~the whole genome background. ORA is
  the first and the most basic method of pathway analysis~\cite{Campos}. It uses a
DEG list with user defined cut-off for the log-fold change and
\textit{p}-value (most commonly using absolute log-fold change $\geq$
2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
associated with annotated pathways are tested for being
over-represented in the set of DEGs. To this end, hyper-geometric
distribution, chi-square tests, binomial probability or the Fisher’s
exact test are used. Thereby the information of the topology of genes
in the pathways is neglected~\cite{Bayer}. Furthermore, ORA assumes
that the biological pathways are independent of each other and ignores
the fact that they cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

In the next step, we check if any of the initial modulated sub-graphs 
could further be extended by adding more than one gene at a time. This is done
by testing all possible extension paths up to \emph{N} (default 2) genes at
all nodes in the sub-graph. Again, this step is iteratively repeated until
no further genes are added to the significant differentially expressed
sub-graphs. This step bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

@@

After detection and extension of the modulated sub-graphs, they are tested
if combined sub-graphs score better than on their own. The merging of the sub-graphs is done by depth first search traversal from the first sub-graph to the second sub-graph. If the two sub-graphs merge
with the connector of at most \emph{N} genes (default 1 gene) and the
combined \textit{p}-value of the merged sub-graph including the bridging
genes in between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

@@

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, three large modulated sub-graphs were identified with 41,
107, and 18 genes were identified, as well as two with less than 4 genes.
The modulated sub-graphs consist predominantly of cytokines, chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23).
IFNB1 and IFNA1 were both identified as two of the possible sources
in one of the sub-graph identified with 18 genes.
Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes (Supplement material 6H).
At 24~hpi eight modulated sub-graphs were
identified with four main sub-graphs consisting
of 27, 101, 134 and 194 genes, others being smaller than 10 genes. IFNB1
and IFNA1 were identified as the two sources out of the total sources. For the last time-point 48~hpi,
six modulated sub-graphs were identified. Three of the sub-graphs were
less than nine genes and main sub-graphs had 122, 191 and 202 genes. IFNB1 was identified as one of the sources in
the most significantly modulated sub-graph (Supplment).

@@

\noindent
Source code: \newline
https://github.com/Modulated-Subgraph-Finder/MSF \newline Software
license: Creative Commons Attribution 4.0 International License

\subsection*{Supplement Material}

\noindent
https://github.com/Modulated-Subgraph-Finder/MSF

@@

\begin{table*}[]
	\centering
	\caption{Comparison of connected sub-graphs of modulated genes
          in the global network between the analysis results of
          \texttt{MSF} and mapping the raw list of differentially
          expressed genes from the standard edgeR analysis, applying
          different p-value cut-offs, onto the network. }
	\label{tab:rawVsHartung}
	\begin{tabular}{lll}
		\hline
		\multicolumn{1}{|l|}{}                            & \multicolumn{1}{c|}{\textbf{Total number of}}       & \multicolumn{1}{c|}{\textbf{Number of connected}}      \\
                \multicolumn{1}{|l|}{}                            & \multicolumn{1}{c|}{\textbf{genes in network}}      & \multicolumn{1}{c|}{\textbf{sub-graph in network}}     \\ \hline
		\multicolumn{1}{|l|}{\textbf{6hpi}}               & \multicolumn{1}{l|}{}                                                & \multicolumn{1}{c|}{}                 \\ \hline
		\multicolumn{1}{|l|}{edgeR + MSF}                 & \multicolumn{1}{c|}{166}                                             & \multicolumn{1}{c|}{3}                \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.1)}  & \multicolumn{1}{c|}{123}                                             & \multicolumn{1}{c|}{39}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.05)} & \multicolumn{1}{c|}{123}                                             & \multicolumn{1}{c|}{47}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.01)} & \multicolumn{1}{c|}{112}                                             & \multicolumn{1}{c|}{53}               \\ \hline
		\multicolumn{1}{|l|}{\textbf{24hpi}}              & \multicolumn{1}{c|}{}                                                & \multicolumn{1}{c|}{}                 \\ \hline
		\multicolumn{1}{|l|}{edgeR + MSF}                 & \multicolumn{1}{c|}{456}                                             & \multicolumn{1}{c|}{4}                \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.1)}  & \multicolumn{1}{c|}{325}                                             & \multicolumn{1}{c|}{76}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.05)} & \multicolumn{1}{c|}{305}                                             & \multicolumn{1}{c|}{102}              \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.01)} & \multicolumn{1}{c|}{262}                                             & \multicolumn{1}{c|}{110}              \\ \hline
		\multicolumn{1}{|l|}{\textbf{48hpi}}              & \multicolumn{1}{c|}{}                                                & \multicolumn{1}{c|}{}                 \\ \hline
		\multicolumn{1}{|l|}{edgeR + MSF}                 & \multicolumn{1}{c|}{515}                                             & \multicolumn{1}{c|}{3}                \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.1)}  & \multicolumn{1}{c|}{392}                                             & \multicolumn{1}{c|}{43}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.05)} & \multicolumn{1}{c|}{366}                                             & \multicolumn{1}{c|}{74}               \\ \hline
		\multicolumn{1}{|l|}{edgeR (p-value $\leq$ 0.01)} & \multicolumn{1}{c|}{306}                                             & \multicolumn{1}{c|}{106}               \\ \hline
	\end{tabular}
\end{table*}
